\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology

### 2022-02-16 jsdse/Manuscript.tex
almost there

\if0\blind
{
  \title{\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge ``Sustaining the Open Tree of Life'',
    NSF ABI No. 1759838, and ABI No. 1759846. They also acknowledge ``The Carpentries''
    project, without it this project could not have been possible}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

@@

All of these resources
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are available for download and use from OpenTree free of cost to any user.
One way to access OpenTree resources is through it Graphical User
Interface (GUI; aka, a website or application that allows users to access computer
functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key
for reproducibility, as manual data manipulation scales poorly and is prone to
error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application
Programming Interface services (APIs; aka, computer code that implements computer functionalities,
in this case OpenTree resources, that can be used by programmers
to build more functionalities),
which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerable more
computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module
\citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
and make them available through the R and Python programming languages.
R and Python are open source and free of cost, and represent
two of the most widely used programming languages in the sciences today \citep{eglen2009quick,
baker2017scientific}.
As such, \texttt{rotl} and \texttt{opentree} software packages should contribute
to making OpenTree resources more accessible to a wider programming user audience.

@@

A third important aspect of software documentation are examples that demonstrate
usage of individual functions \citep{karimzadeh2018top}.
Examples presented in software documentation are usually
worked to perfection, as they are intended to showcase the ideal or minimal case
in which a function works well.
Perfectly worked examples ignore the user experience by
maintaining focus on the software content, and fail to provide users with expert
and clear advice on how to troubleshoot if needed.
We identify this as \textit{barrier 3.-- Content-focused examples}.
Error management training is an approach that focuses on framing mistakes as beneficial
to learning complex tasks, to give learners the opportunity to
actively explore a task \citep{frese1995error}.
Providing examples that showcase potential errors supports users performance \citep{steele2014error},
and can greatly improve learners ability to troubleshoot outside the classroom
\citep{shannon2015live, nederbragt2020ten}.

@@

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (see Figure \ref{fig:error}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as
individual users to identify such specialized concepts.

@@

Cognitive load can be greatly reduced for learners by applying an active learning
strategy such as linking usage to a ``real world'' or ``human'' application
\citep{felder2009active}.
Programming computer languages are by themselves quite abstract and represent a
learning subject with a potentially high cognitive load for most learners.
Pedagogical research shows that active learning practices are one of the most effective
ways to take on abstract subjects \citep{freeman2014active}.

A story-like narrative that links code usage in an integrative example, invites learners
to try the code, which can lead them to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following
the code usage, supporting memory and understanding \citep{piccolo2016tools}.

We propose that documents developed with ``literate programming'' can be made more
accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as model organism.
For a research group this can be the group they are studying. For the general audience,
a highly charismatic group such as dinosaurs will do the trick.

We examined available primary documentation for the package \texttt{rotl},
and designed a narrative that required the usage of as many functions as possible.
We demonstrate code applications that are commonly requested by OpenTree users,
but that are not demonstrated in the primary documentation of the R package.
By framing the function workflow using highly requested uses, the documentation acquires a
narrative arc that is easier to follow and remember by users. This can also facilitate
the application of code to other use cases in biology of interest for the users.

%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested
%% use case of obtaining
%% a phylogenetic tree for all lineages within a specific taxonomic rank.

@@

A practice that has become more and more widespread in programming-language pedagogical practices
is live programming.
One benefit of live programming is that typos and mistakes occur -- which normalizes
them for learners, and shows them how to solve them when they are outside the
classroom \citep{shannon2015live, nederbragt2020ten}.
When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer
with the Carpentries (a non-profit group that teaches foundational coding and
 data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
a senior faculty member taking the workshop
complained that the typos were slowing things down and interfering with the pedagogy.
McTavish replied "the typos ARE the pedagogy".
This has become a slogan of sorts at the Carpentries, capturing the idea that embracing
and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
Yet, working through mistakes is rarely done on written pedagogical materials.
Software documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
and users performance
\citep{gaspar2007restoring, steele2014error}.
In our tutorials, we apply this principle by demonstrating examples that do not work
as expected, and exemplifying ways to address them (Figure \ref{fig:error}).

We identified inputs that would give
a wide range of warnings and errors, and support learners on reading the information
provided by the message. This
helps users to not be afraid of errors and warnings, but instead use them as
an informative measure.
% This is probably part of the computer science curriculum, but it is not at all part of the natural sciences curriculum.
We also identify effects of warnings and errors downstream of the workflow.

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.

In our tutorial, we focus on explaining the meaning of warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and errors, and
2) it allows users/students to understand with more depth what the function is doing.

@@

Making approachable research workflows has several advantages:

* save explanation/training time when analyses are run again by students and collaborators;
* save research time for yourself when analyses are run again with more data, a* different dataset, a different organism or biological model;
* scientific efforts can build off of each other.

@@

The principles to create tutorials described here facilitate adoption of software
 and analysis workflows among researchers at different academic levels, from undergrads
  to established researchers.
It will also help close the gap between students that had access to computational
resources (and computational training) from an early age and students that did not.
Late access to computational resources and training can occur due to lack of
economic resources, often occurring in households from underrepresented communities
and minorities \citep{google2016diversity, warner2021quantifying}.
It can also be due to gender-biased parental and community pressures,
in which male individuals are more often encouraged to perform activities related to computers,
while female individuals are discouraged \citep{warner2021quantifying}.
% How to balance software acceptance VS. adoption?
These principles can be used to improve not only reproducibility practices,
 but also software adoption in the natural sciences.
%% Discuss: why address accessibility and not other aspects of reproducibility?

@@

\bigskip
\bigskip
\bigskip

@@

@@

>>>>

\if0\blind
{
  \title{\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge ``Sustaining the Open Tree of Life'',
    NSF ABI No. 1759838, and ABI No. 1759846. They are also also deeply grateful towards
    ``The Carpentries'' organization; without its invaluable volunteers and resources
    this project could not have been possible}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

@@

All of these resources
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are available for download and use from OpenTree free of cost to any user.
One way to access OpenTree resources is through its Graphical User
Interface (GUI; aka, a website or application that allows users to access computer
functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key
for reproducibility, as manual data manipulation scales poorly and is prone to
error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application
Programming Interface services (APIs; aka, computer code that implements computer functionalities,
in this case OpenTree resources, that can be used by programmers
to build more functionalities),
which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerable more
computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module
\citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
and make them available through the R and Python programming languages.
R and Python are open source and free of cost, and represent
two of the most widely used programming languages in the sciences today \citep{eglen2009quick,
baker2017scientific}.
As such, \texttt{rotl} and \texttt{opentree} software packages should contribute
to making OpenTree resources more accessible to a wider programming user audience.

@@

A third important aspect of software documentation are examples that demonstrate
usage of individual functions \citep{karimzadeh2018top}.
Examples presented in software documentation are usually
worked to perfection, as they are intended to showcase the ideal or minimal case
in which a function works well.
Perfectly worked examples ignore the user experience by
maintaining focus on the software content, and fail to provide users with expert
and clear advice on how to troubleshoot if needed.
We identify this as \textit{barrier 3.-- Content-focused examples}.
Error management training is an approach that focuses on framing mistakes as beneficial
to learning complex tasks, to give learners the opportunity to
actively explore a task with a positive mindset \citep{frese1995error}.
Providing examples that showcase potential errors supports users performance \citep{steele2014error},
and can greatly improve learners ability to troubleshoot outside the classroom
\citep{shannon2015live, nederbragt2020ten}.

@@

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (Figure \ref{fig:error}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as
individual users to identify such specialized concepts.

@@

Cognitive load can be greatly reduced for learners by applying an active learning
strategy such as linking usage to a ``real world'' or ``human'' application
\citep{felder2009active}.
Programming computer languages are by themselves quite abstract and represent a
learning subject with a potentially high cognitive load for most learners.
Pedagogical research shows that active learning practices are one of the most effective
ways to take on abstract subjects \citep{freeman2014active}.
A story-like narrative that links code usage in an integrative example, invites learners
to try the code, which can lead them to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following
the code usage, supporting memory and understanding \citep{piccolo2016tools}.

We propose that documents developed with ``literate programming'' can be made more
accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as model organism.
For a research group this can be the group they are studying. For the general audience,
a highly charismatic group such as dinosaurs will do the trick.
For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb}
in January 2020, we tailored all examples to frogs and their allies.
% \citep{cite-aw-workshop}

We examined available software documentation for the package \texttt{rotl},
and designed a narrative that requires the usage of as many functions as possible.
We demonstrate code applications that are commonly requested by OpenTree users,
but that are not demonstrated in the documentation of the package.
By framing the function workflow using highly requested uses, the documentation acquires a
narrative arc that is easier to follow and remember by users. This can also facilitate
translating the code application to other use cases of interest for learners in biology.

%% Maybe Figure!! In the tutorial showcased here, we used the commonly requested
%% use case of obtaining
%% a phylogenetic tree for all lineages within a specific taxonomic rank.

@@

A practice that has become more and more widespread in programming-language pedagogical practices
is live programming.
One benefit of live programming is that typos and mistakes occur -- which normalizes
them for learners, and shows them how to solve them when they are outside the
classroom \citep{shannon2015live, nederbragt2020ten}.
When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer
with the Carpentries (a non-profit group that teaches foundational coding and
 data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
a senior faculty member taking the workshop
complained that the typos were slowing things down and interfering with the pedagogy.
McTavish replied "the typos ARE the pedagogy".
This has become a slogan of sorts at the Carpentries, capturing the idea that embracing
and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
Yet, working through mistakes is rarely done on written pedagogical materials.
Software documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
and users performance \citep{gaspar2007restoring, steele2014error}.

In our tutorials, we apply this principle by demonstrating examples that do not work
as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
For example, we identified inputs that would give
a wide range of warnings and errors. We then focus on providing explanations for
error and warning messages. We believe this supports users and learners to be less
intimidated by the messages and practice taking useful information out of it.
% This is probably part of the computer science curriculum, but it is not at all
% part of the natural sciences curriculum.

We also demonstrate ways to evaluate inputs to determine if they will trigger an
error or warning, and design and demonstrate alternative analysis routes on what
to do when faced with an error or warning.
One of the most essential skills in programming is interpreting and moving forward
from errors.
This has two pedagogical benefits. First, it provides users and learners with the
means to troubleshoot their own warnings and errors. Second, it allows them to
understand with more depth what the function is doing.

@@

Making approachable research workflows has all the advantages of reproducible research
workflows.
It saves explanation/training time when analyses are run again by students and
collaborators.
It can also save research time for yourself when analyses are run again with more
data, a different dataset, a different organism or biological model;
It contributes to scientific efforts that can build off of each other.

@@

The principles to create tutorials described here facilitate adoption of software
and analysis workflows among researchers at different academic levels, from undergrads
to established researchers.
It will also help close the gap between students that had access to computational
resources (and computational training) from an early age and students that did not.
Late access to computational resources and training can occur due to lack of
economic resources, often occurring in households from underrepresented communities
and minorities \citep{google2016diversity, warner2021quantifying}.
It can also be due to gender-biased parental and community pressures,
in which male individuals are more often encouraged to perform activities related to computers,
while female individuals are discouraged \citep{warner2021quantifying}.
These principles can be used to improve not only reproducibility practices,
but also software adoption in the natural sciences.

@@

\pagebreak

@@

\newline
\newline
\newline

@@

\pagebreak
